### 2018-07-03 paper/4-build.tex
Comments from Simon

Section \S\ref{sec-background} introduced three different \emph{task scehedulers}
that decide which tasks to execute and in what order; see the ``Scheduler'' column
of Table~\ref{tab-summary} in \S\ref{sec-background-summary}.
This subsection explores the properties of the three schedulers, and
possible implementations.

@@

The topological scheduler has two downsides: it is limited to \hs{Applicative}
build systems and requires a fresh topological sort each time. So, while the
actions themselves may be incremental (i.e. unnecessary tasks will not be
performed), the pre-processing is not. We can incrementalise the topological
sort by storing the topological order between build runs, using it as a starting
point in the next run, fixing it up if the build discovers any inconsistency, as
\Excel does.

@@

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it, and
also making sure all cells that depend on it are recursively marked dirty too.
When rebuilding, if a clean cell only depends on clean cells it is skipped,
otherwise it is rebuilt.

@@

\Make uses file modification times, and compares files to their
dependencies, which can be thought of as a dirty bit which is set when
a file is newer than its dependencies. The interesting property of
this dirty bit is that it is not under the control of \Make; rather it is
existing file-system information that has been repurposed. In particular,
modifying a file automatically clears its dirty bit, and
automatically sets the dirty bit of the nodes depending on it. One
thing \Make does require is that file timestamps only go forward in
time -- something that can be violated by backup software.

@@

With a dirty bit it is simple to achieve minimality. However, to achieve early
cutoff (\S\ref{sec-background-shake}) it would be important to not set the dirty
bit after a computation that did not change the value. \Excel could use this
approach, but does not. In contrast, \Make cannot implement early cutoff nicely
-- to do so it would have to mark the key clean (so it would not rebuild in the next
run) and at the same time not mark the keys that depend on it dirty -- an
impossible task with only the ability to update to the latest modification time.
\Make can approximate early cutoff by not modifying the result file, and not
marking it clean, but then it will rerun in every subsequent build.

@@

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well if we skip any
number $n$ of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$.

>>>>

Section \S\ref{sec-background} introduced three different \emph{task schedulers}
that decide which tasks to execute and in what order; see the ``Scheduler'' column
of Table~\ref{tab-summary} in \S\ref{sec-background-summary}.
This subsection explores the properties of the three schedulers, and
possible implementations.

@@

The topological scheduler has two downsides: it is limited to \hs{Applicative}
build systems and requires a fresh topological sort each time. \simon{If the tasks don't change,
  the dependencies are unchanged, so the topological sort from the last time
  will do fine, won't it?}  So, while the
actions themselves may be incremental (i.e. unnecessary tasks will not be
performed), the pre-processing is not. We can incrementalise the topological
sort by storing the topological order between build runs, using it as a starting
point in the next run, fixing it up if the build discovers any inconsistency, as
\Excel does.

@@

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it, and
also making sure all cells that depend on it are recursively marked dirty too.
\simon{Not always (see make) and only possible if we have static dependencies
  or have recorded dynamic dependencies.}
When rebuilding, if a clean cell only depends on clean cells it is skipped,
otherwise it is rebuilt. \simon{But if all the dependencies of a dirty cell
  are marked dirty, we rebuild only dirty cells not clean ones, contradicting this statement.}

@@

\Make uses file modification times, and compares files to their
dependencies, which can be thought of as a dirty bit which is set when
a file is newer than its dependencies. The interesting property of
this dirty bit is that it is not under the control of \Make; rather it is
existing file-system information that has been repurposed. In particular,
modifying a file automatically clears its dirty bit, and
automatically sets the dirty bit of the nodes depending on it. One
thing \Make does require is that file timestamps only go forward in
time -- something that can be violated by backup software. \simon{Also transitive dependencies
  of dirty files are not made dirty.}

@@

With a dirty bit it is simple to achieve minimality. However, to achieve early
cutoff (\S\ref{sec-background-shake}) it would be important to not set the dirty
bit after a computation that did not change the value. \simon{And if we'd recursively
made the transitive dependencies dirty, we could not do early cutoff.}
\Excel could use this
approach, but does not. \simon{It'd be tricky becase of setting dirtiness recursively.}
In contrast, \Make cannot implement early cutoff nicely
-- to do so it would have to mark the key clean (so it would not rebuild in the next
run) and at the same time not mark the keys that depend on it dirty -- an
impossible task with only the ability to update to the latest modification time.
\Make can approximate early cutoff by not modifying the result file, and not
marking it clean, but then it will rerun in every subsequent build.

@@

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well if we skip any
number of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$.

### 2018-07-03 paper/5-implementations.tex
Comments from Simon

@@

\Make's scheduler, \hs{topological}, processes keys in a linear \hs{order} based on a
topological sort of the statically known dependency graph
(see~\S\ref{sec-parallelism} for parallel \Make). Our definition in
Fig.~\ref{fig-make-implementation} is polymorphic with respect to the type of
build information \hs{i} and is therefore compatible with any applicative
\hs{rebuilder}. The scheduler calls the supplied \hs{rebuilder} on every
\hs{key} in the \hs{order}, and runs the obtained \hs{newTask} to compute the
\hs{newValue}. Note that \hs{newTask} has access only to the \hs{i} part of the
\hs{Store}~\hs{i}~\hs{k}~\hs{v}, but the rest of the \hs{do} block runs in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad; we use the (unremarkable)
helper function \hs{liftStore} to fix the mismatch. The \hs{newTask} finds
values of the \hs{key}'s dependencies via the \hs{fetch} callback, which is
defined to directly read the \hs{store}.

@@

Our model of \Excel uses the \hs{restarting} scheduler and the
\hs{dirtyBitRebuilder}, see Fig.~\ref{fig-excel-implementation}. The persistent
build information \hs{ExcelInfo}~\hs{k} is a pair: (i) a map
\hs{k}~\hs{->}~\hs{Bool} associating a dirty bit with every key, and (ii) a
calc \hs{Chain}~\hs{::}~\hs{[@@k]} recorded in the previous build
(\S\ref{sec-background-excel}).

@@

Using the abstractions built thus far, we have shown how to combine schedulers
with rebuilders to reproduce existing build systems. To us, the most interesting
build system as yet unavailable would combine a suspending scheduler with
constructive traces~--~providing a cloud-capable build system that is minimal,
and supports both early cutoff and monadic dependencies. Using our framework it
is possible to define and test such a system, which we call \Cloud \Shake. All
we need to do is replace \hs{restarting2} with \hs{suspending} in the definition
of \Bazel, as shown in Fig.~\ref{fig-cloud-implementations}.

>>>>

The scheduler calls the rebuilder, but passes to it a \hs{fetch} function
that that latter calls when it needs the value of a dependent key.  This
callback returns control to the scheduler, which may in turn call the
rebuilder to bring the dependent key up to date, and so on.

@@

\simon{topological is hard to read because of the execState and forM stuff.
  Could we have a local defn, with a type signature, for the $\lambda key. blah$?
  If necessary we could omit the code for liftStore, leaving only its signature}
\Make's scheduler, \hs{topological}, processes keys in a linear \hs{order} based on a
topological sort of the statically known dependency graph
(see~\S\ref{sec-parallelism} for parallel \Make). Our definition in
Fig.~\ref{fig-make-implementation} is polymorphic with respect to the type of
build information \hs{i} and is therefore compatible with any applicative
\hs{rebuilder}. The scheduler calls the supplied \hs{rebuilder} on every
\hs{key} in the \hs{order}, and runs the obtained \hs{newTask} to compute the
\hs{newValue}. Note that \hs{newTask} has access only to the \hs{i} part of the
\hs{Store}~\hs{i}~\hs{k}~\hs{v}, but the rest of the \hs{do} block runs in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad; we use the (unremarkable)
helper function \hs{liftStore} to fix the mismatch. The \hs{newTask} finds
values of the \hs{key}'s dependencies via the \hs{fetch} callback, which is
defined to directly read the \hs{store}.

@@

Our model of \Excel uses the \hs{restarting} scheduler and the
\hs{dirtyBitRebuilder}, see Fig.~\ref{fig-excel-implementation}. The persistent
build information \hs{ExcelInfo}~\hs{k} is a pair: (i) a map
\hs{k}~\hs{->}~\hs{Bool} associating a dirty bit with every key, and (ii) a
calc chain of type \hs{[@@k]} recorded in the previous build
(\S\ref{sec-background-excel}).

@@

\simon{New subsection here? Or at least put this para after the next one.}
Using the abstractions built thus far, we have shown how to combine schedulers
with rebuilders to reproduce existing build systems. To us, the most interesting
build system as yet unavailable would combine a suspending scheduler with
constructive traces~--~providing a cloud-capable build system that is minimal,
and supports both early cutoff and monadic dependencies. Using our framework it
is possible to define and test such a system, which we call \Cloud \Shake. All
we need to do is replace \hs{restarting2} with \hs{suspending} in the definition
of \Bazel, as shown in Fig.~\ref{fig-cloud-implementations}.
